---
title: "WGCNA Analysis"
author: "Asad"
date: "2023-05-12"
output: 
  html_document: 
    highlight: zenburn
    theme: journal
---
### Call libraries

```{r}
library(WGCNA)
library(DESeq2)
library(tidyverse)
library(CorLevelPlot)
library(gridExtra)
library(magrittr)
library(ggplot2)

allowWGCNAThreads()

setwd('E:/WGCNA')
```

### 1. Read data
```{r}
data<- read.csv('Counts.csv', row.names = 1)
Sample_info<- read.csv('Sample_info.csv', row.names = 1)
```

### 2. Quality control
#### Step1: Detect outlier genes with WGCNA package
```{r}
gg<- goodSamplesGenes(t(data))
summary(gg)
gg$allOK
```

#Since it indicates tat all genes are not good genes let's quantify the outlier genes
#We'll do it for samples as well
```{r}
table(gg$goodGenes)
table(gg$goodSamples)
```
#Around 3500 outlier genes but no outlier samples. Let's remove those genes
```{r}
data<- data[gg$goodGenes=='TRUE',]
```

#### Step2: Identify outlier samples with hierachical clustering
```{r}
htree <- hclust(dist(t(data)), method = "average")
plot(htree)
```
#### Step3: Identify outlier samples with PCA
```{r}
pca <- prcomp(t(data))
pca.dat <- pca$x

pca.var <- pca$sdev^2
pca.var.percent <- round(pca.var/sum(pca.var)*100, digits = 2)

pca.dat <- as.data.frame(pca.dat)

ggplot(pca.dat, aes(PC1, PC2)) +
  geom_point() +
  geom_text(label = rownames(pca.dat)) +
  labs(x = paste0('PC1: ', pca.var.percent[1], ' %'),
       
       y = paste0('PC2: ', pca.var.percent[2], ' %'))

#N.B. We need to remove batch effect if there's any at this stage before we proceed further

#In both these methods we found Sample5 is distantly related 
#So we'll exclude them
discard_samples<- c('Sample5')
data.final<- data[,!colnames(data) %in% discard_samples]
Sample_info_final<- Sample_info[!row.names(Sample_info) %in% discard_samples,]

```

### 4. Perform VST normalization with DESeq2 since it's a count matrix
#N.B. FPKM or RPKM matrices need to be log transformed
#Before performing normalization make sure if the rownames in matrix and colnames in
#Sample info are identical
```{r}
all(rownames(Sample_info_final)%in%colnames(data.final)) #If present
all(rownames(Sample_info_final)==colnames(data.final)) #If present in same order
```

#### Create deseq2 object
```{r}
dds <- DESeqDataSetFromMatrix(countData = data.final,
                              colData = Sample_info_final,
                              design = ~ 1) #Specifying no design
nrow(dds)
```

#### Remove samples that have <15 counts in more than 75% of the samples
#Recommended by WGCNA
#However, the downstream analysis with higher # of genes  will depend largely on 
#the available ram of PC. So, we'll consider gene numbers based on our PC capacity.
```{r}
keep <- dds[rowSums(counts(dds) >= 15)>=39,] #75% of 52 samples is 52*0.75=39
nrow(keep)
```

#### Perform VST transformation
```{r}
dds.vst<- vst(keep) 
dds.vst<- dds.vst[1:5000,] #We reduced the # of genes to 8000 because of low RAM
nrow(dds.vst)
```

# Get normalized counts
```{r}
norm.counts <- assay(dds.vst) %>% 
  t()
```

### 5. Constructing network
# Set softhreshold power
```{r}
power <- c(c(1:10), seq(from = 12, to = 50, by = 2))
```

# Set the network topology analysis function
```{r}
sft <- pickSoftThreshold(norm.counts,
                         powerVector = power,
                         networkType = "signed",
                         verbose = 5)
```

#N.B. unsigned -> nodes with positive & negative correlation are treated equally 
#N.B. signed -> nodes with negative correlation are considered *unconnected*, treated as zero
```{r}
sft.data <- sft$fitIndices
```
# visualization to pick power

```{r}
Plot1 <- ggplot(sft.data, aes(Power, SFT.R.sq, label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.8, color = 'red') +
  labs(x = 'Power', y = 'Scale free topology model fit, signed R^2') +
  theme_classic()


Plot2 <- ggplot(sft.data, aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  labs(x = 'Power', y = 'Mean Connectivity') +
  theme_classic()


grid.arrange(Plot1, Plot2, nrow = 2)
```

# convert matrix to numeric and assign power
```{r}
norm.counts[] <- sapply(norm.counts, as.numeric)
soft_power <- 14
temp_cor <- cor
cor <- WGCNA::cor
```

# Estimate w.r.t blocksize memory depends on available ram of PC
```{r}
bwnet <- blockwiseModules(norm.counts,
                          maxBlockSize = 5000,
                          TOMType = "signed",
                          power = soft_power,
                          mergeCutHeight = 0.25,
                          numericLabels = FALSE,
                          randomSeed = 1234,
                          verbose = 3)
cor <- temp_cor
```

### 6. Identify Module Eigengenes 
```{r}
module_eigengenes <- bwnet$MEs
head(module_eigengenes)
```
# Check the number of genes for each module
```{r}
table(bwnet$colors)
```
# Plot the dendrogram and the module colors before and after merging underneath
```{r}
plotDendroAndColors(bwnet$dendrograms[[1]], cbind(bwnet$unmergedColors, bwnet$colors),
                    c("unmerged", "merged"),
                    dendroLabels = FALSE,
                    addGuide = TRUE,
                    hang= 0.03,
                    guideHang = 0.05)
```

# grey module = all genes that doesn't fall into other modules were assigned to the grey module


### 7. Extract genes and their associated color modules and save them
```{r}
mergedColors<- labels2colors(bwnet$colors)
module_df <- data.frame(
  gene_id <- names(bwnet$colors),
  colors <- labels2colors(bwnet$colors)
)

module_df[1:5,]

write_delim(module_df,
            file = "gene_modules.txt",
            delim = "\t")
```

### 8. Merging and clustering modules from eigengenes
# Calculate eigengenes
```{r}
MEList <- moduleEigengenes(norm.counts, colors = mergedColors)
MEs <- MEList$eigengenes
```

# Calculate dissimilarity of module eigengenes
```{r}
MEDiss <- 1-cor(MEs)
METree <- hclust(as.dist(MEDiss), method = "average")
```
# Plot the result
```{r}
plot(METree, main = "Clustering of Module eigengenes",
     xlab = "", sub = "")
```

### 9. Show the correlation between modules and each sample
# Get Module Eigengenes per cluster
```{r}
MEs0 <- moduleEigengenes(norm.counts, mergedColors)$eigengenes
```

#Reorder modules so similar modules are next to each other
```{r}
MEs0 <- orderMEs(MEs0)
module_order <- names(MEs0) %>% gsub("ME","", .)
```

# Add the treatment names
```{r}
MEs0$treatment <- row.names(MEs0)
```

# Tidy & plot data
```{r}
mME = MEs0 %>%
  pivot_longer(-treatment) %>%
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order) 
  )

mME %>% ggplot(., aes(x=treatment, y=name, fill=value)) +
  geom_tile() +
  theme_bw() + 
  scale_fill_gradient2(
    low = "blue", 
    high = "red", 
    mid = "white", 
    midpoint = 0, 
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")
```

### 10. Showing gene expression level in specific modules
```{r}

modules_of_interest = c("green", "turquoise", "red")
```

# Pull out list of genes in that module
```{r}
submod = module_df %>%
  subset(colors %in% modules_of_interest)

row.names(module_df) = module_df$gene_id
```

# Get normalized expression for those genes
```{r}
norm.counts[1:5,1:10]
norm.exprs<- data.frame(t(norm.counts)) 
norm.exprs$gene_id<- row.names(norm.exprs) 
norm.exprs<- norm.exprs %>% select(gene_id,everything())
subexpr <- norm.exprs[submod$gene_id,]
 
submod_df <- data.frame(subexpr) %>%
  mutate(
    gene_id = row.names(.)
  ) %>% 
  pivot_longer(-gene_id) %>%
  mutate(
    module = module_df[gene_id,]$colors
  )

submod_df %>% ggplot(., aes(x=name, y=value, group=gene_id, colour=module)) +
  geom_line(alpha=0.2) + 
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  facet_grid(rows = vars(module)) +
  labs(x = "Treatment",
       y = "Normalized expression") + 
  scale_color_manual(values = c("lightgreen","red","turquoise"))
```
### 11. Finding the association between module and traits
# Prepare and manipulate data for association analysis
# Converting categorical variable into binary group variables
```{r}
type <- Sample_info_final %>% 
        mutate(type = ifelse(grepl('Lung cancer', Group), 1, 0)) %>% 
        select(4)
```

```{r}
Sample_info_final$Status <- factor(Sample_info_final$Status, 
                                   levels = c("None", "Newly diagnosed", 
                                   "Post Chemotherapy", "On Doxorubicin"))
```


```{r}
Status<- binarizeCategoricalColumns(Sample_info_final$Status,
                                           includePairwise = FALSE,
                                           includeLevelVsAll = TRUE,
                                           minCount = 1)


Status.final<- cbind(type, Status)
```

# Define numbers of genes and samples and draw correlation
```{r}
nSamples <- nrow(norm.counts)
nGenes <- ncol(norm.counts)
status.module.corr <- cor(module_eigengenes, Status.final, use = 'p')
status.module.corr.pvals <- corPvalueStudent(status.module.corr , nSamples)
nrow(module_eigengenes)
```

####  Visualize module-trait association as a heatmap
```{r}
heatmap.data <- merge(module_eigengenes, Status.final, by = 'row.names')

head(heatmap.data)

heatmap.data <- heatmap.data %>% 
   column_to_rownames(var = 'Row.names')

CorLevelPlot(heatmap.data,
             x = names(heatmap.data)[9:12],
             y = names(heatmap.data)[1:11],
             col = c("blue1", "green", "pink", "red"))



module.gene.mapping <- as.data.frame(bwnet$colors)
module.gene.mapping %>% 
  filter(`bwnet$colors` == 'blue') %>% 
  rownames()

```

### 12. Identifying significant genes against different features of data
#### Calculating the module membership and the associated p-values
# The module membership/intramodular connectivity is calculated as the correlation of the eigengene and the gene expression profile. 
# This quantifies the similarity of all genes on the array to every module.
```{r}
module.membership.measure <- cor(module_eigengenes, norm.counts, use = 'p')
module.membership.measure.pvals <- corPvalueStudent(module.membership.measure, nSamples)

module.membership.measure.pvals[1:5,1:10]
```

#### Calculating the gene significance and associated p-values against a particular trait
```{r}
gene.signf.corr <- cor(norm.counts, Status.final$'data.Post Chemotherapy.vs.all', use = 'p')
gene.signf.corr.pvals <- corPvalueStudent(gene.signf.corr, nSamples)
```
# Sub-setting based on significant association: X1:corr, X2:pval
```{r}
Sig_chemo<-data.frame(cbind(gene.signf.corr, gene.signf.corr.pvals))
Sig_chemo_final<- filter(Sig_chemo, Sig_chemo$X1> 0.3 | Sig_chemo$X1< -0.3
                         & Sig_chemo$X2<0.05)  

write.csv(Sig_chemo_final, 'sig_genes_chemo.csv')
```






